---
title: "cs10: Satellite Remote Sensing"
format: html
editor: visual
---

# **Overview**

## **1. Introduction**

In this case study, you’ll explore how satellite remote sensing can reveal spatial and temporal patterns in **Land Surface Temperature (LST)** and **Land Cover (LULC)** using MODIS data products.

By the end of this lab, you will: - Read and interpret data stored in **NetCDF** format\
- Extract and analyze time series from satellite data\
- Calculate **monthly climatologies**\
- Compare **LST across land cover types**

You’ll work primarily with two MODIS datasets: - **MOD11A2**: Land Surface Temperature (1 km, 8-day composite) - **MCD12Q1**: Land Cover Type (annual classification)

# **Setup and Data Download**

You’ll need both `MOD11A2` (LST) and `MCD12Q1` (Land Cover) NetCDF files.

### **💡 Hints:**

-   Load key packages: `terra`, `tidyverse`, `sf`, and `ncdf4`

-   Create a `data` folder for downloads using `dir.create("data", showWarnings = FALSE)`

-   Download the `.nc` files from the provided GitHub URLs

-   Store them in a `data/` folder and, if you are keeping your code in a git repository, add `*data*` to your `.gitignore` file so git doesn’t track them.

```{r}
#install.packages("rasterVis")
#install.packages("ncdf4")
library(terra)
library(tidyverse)
library(sf)
library(knitr)
library(kableExtra)
library(rasterVis)
library(ncdf4)

dir.create("data", showWarnings = FALSE)

lulc_url <- "https://github.com/adammwilson/DataScienceData/blob/master/inst/extdata/appeears/MCD12Q1.051_aid0001.nc?raw=true"
lst_url <- "https://github.com/adammwilson/DataScienceData/blob/master/inst/extdata/appeears/MOD11A2.006_aid0001.nc?raw=true"

download.file(lulc_url, destfile = "data/MCD12Q1.051_aid0001.nc", mode = "wb")
download.file(lst_url, destfile = "data/MOD11A2.006_aid0001.nc", mode = "wb")
```
Load and Explore the Data
💡 Hints:

Use nc_open() to print the file metadata and learn the names of the subdataset
Use rast() to read the NetCDF file and specify which subdatasets. (sbds) you want to select.
Check layer names with names().
Use plot() or gplot() to visualize the rasters.

```{r}
MCD12Q1 <- nc_open("data/MCD12Q1.051_aid0001.nc")
MOD11A2 <- nc_open("data/MOD11A2.006_aid0001.nc")

lulc <- rast("data/MCD12Q1.051_aid0001.nc", subds = "Land_Cover_Type_1")

names(lulc)

plot(lulc$Land_Cover_Type_1_13)

lst <- rast("data/MOD11A2.006_aid0001.nc", subds ="LST_Day_1km")

names(lst)

plot(lst)
```
# **Convert and Label Land Cover**

```{r}

Land_Cover_Type_1 <- c(
  "Water" = 0, "Evergreen Needleleaf forest" = 1, "Evergreen Broadleaf forest" = 2,
  "Deciduous Needleleaf forest" = 3, "Deciduous Broadleaf forest" = 4, "Mixed forest" = 5,
  "Closed shrublands" = 6, "Open shrublands" = 7, "Woody savannas" = 8, "Savannas" = 9,
  "Grasslands" = 10, "Permanent wetlands" = 11, "Croplands" = 12, "Urban & built-up" = 13,
  "Cropland/Natural vegetation mosaic" = 14, "Snow & ice" = 15, "Barren/Sparsely vegetated" = 16
)

lcd <- data.frame(
  ID = Land_Cover_Type_1,
  landcover = names(Land_Cover_Type_1),
  stringsAsFactors = FALSE
)

lulc <- as.factor(lulc)

scoff(lst) = cbind(0.02, -273.15)

plot(lst)

```






