---
title: "cs10: Satellite Remote Sensing"
format: html
editor: visual
---

# **Overview**

## **1. Introduction**

In this case study, you’ll explore how satellite remote sensing can reveal spatial and temporal patterns in **Land Surface Temperature (LST)** and **Land Cover (LULC)** using MODIS data products.

By the end of this lab, you will: - Read and interpret data stored in **NetCDF** format\
- Extract and analyze time series from satellite data\
- Calculate **monthly climatologies**\
- Compare **LST across land cover types**

You’ll work primarily with two MODIS datasets: - **MOD11A2**: Land Surface Temperature (1 km, 8-day composite) - **MCD12Q1**: Land Cover Type (annual classification)

# **Setup and Data Download**

You’ll need both `MOD11A2` (LST) and `MCD12Q1` (Land Cover) NetCDF files.

### **💡 Hints:**

-   Load key packages: `terra`, `tidyverse`, `sf`, and `ncdf4`

-   Create a `data` folder for downloads using `dir.create("data", showWarnings = FALSE)`

-   Download the `.nc` files from the provided GitHub URLs

-   Store them in a `data/` folder and, if you are keeping your code in a git repository, add `*data*` to your `.gitignore` file so git doesn’t track them.

```{r}
#install.packages("rasterVis")
#install.packages("ncdf4")
#install.packages("tidyterra")
library(terra)
library(tidyverse)
library(sf)
library(knitr)
library(kableExtra)
library(rasterVis)
library(ncdf4)
library(lubridate)
library(tidyterra)

dir.create("data", showWarnings = FALSE)

lulc_url <- "https://github.com/adammwilson/DataScienceData/blob/master/inst/extdata/appeears/MCD12Q1.051_aid0001.nc?raw=true"
lst_url <- "https://github.com/adammwilson/DataScienceData/blob/master/inst/extdata/appeears/MOD11A2.006_aid0001.nc?raw=true"

download.file(lulc_url, destfile = "data/MCD12Q1.051_aid0001.nc", mode = "wb")
download.file(lst_url, destfile = "data/MOD11A2.006_aid0001.nc", mode = "wb")
```
Load and Explore the Data
💡 Hints:

Use nc_open() to print the file metadata and learn the names of the subdataset
Use rast() to read the NetCDF file and specify which subdatasets. (sbds) you want to select.
Check layer names with names().
Use plot() or gplot() to visualize the rasters.

```{r}
MCD12Q1 <- nc_open("data/MCD12Q1.051_aid0001.nc")
MOD11A2 <- nc_open("data/MOD11A2.006_aid0001.nc")

lulc <- rast("data/MCD12Q1.051_aid0001.nc", subds = "Land_Cover_Type_1")

names(lulc)

plot(lulc$Land_Cover_Type_1_13, main = "MODIS Land Cover (Year 13)")

lst <- rast("data/MOD11A2.006_aid0001.nc", subds ="LST_Day_1km")

names(lst)
```
# **Convert and Label Land Cover**

```{r}

Land_Cover_Type_1 <- c(
  "Water" = 0, "Evergreen Needleleaf forest" = 1, "Evergreen Broadleaf forest" = 2,
  "Deciduous Needleleaf forest" = 3, "Deciduous Broadleaf forest" = 4, "Mixed forest" = 5,
  "Closed shrublands" = 6, "Open shrublands" = 7, "Woody savannas" = 8, "Savannas" = 9,
  "Grasslands" = 10, "Permanent wetlands" = 11, "Croplands" = 12, "Urban & built-up" = 13,
  "Cropland/Natural vegetation mosaic" = 14, "Snow & ice" = 15, "Barren/Sparsely vegetated" = 16
)

lcd <- data.frame(
  ID = Land_Cover_Type_1,
  landcover = names(Land_Cover_Type_1),
  stringsAsFactors = FALSE
)

lulc <- as.factor(lulc)

scoff(lst) = cbind(0.02, -273.15)

plot(lst, main = "Surface Temperature (C)")

```
Part 1: Extract Time Series for a Point
Hint
Use extract() to get the average LST for a location in Buffalo, NY.

💡 Hints:

Create a point with st_as_sf() to create a point at lon = -78.79, lat = 43.00 in wgs84 (geographic) coordinates (crs = 4326)
Transform to the raster’s CRS with st_transform().
Use terra::extract() to extract LST values within a 1 km buffer.
Combine with the corresponding time(lst) values into a tidy data frame.
Plot the result with ggplot().
`geom_smooth()` using method = 'loess' and formula = 'y ~ x'

```{r}

#Created point to find data around
buff_point <- data.frame(lon = -78.79, lat = 43.00)

#Converted point to sf
buff_point_sf <- st_as_sf(buff_point, coords = c("lon", "lat"), crs = 4326)

#Put sf and raster in the same crs
buff_point_sf <- st_transform(buff_point_sf, st_crs(lst))

#Extracted lst points found around the buff_point_sf
extracted_lst <- terra::extract(lst, buff_point_sf)

#Got times of lst
time_lst <- time(lst)

#Added times to the extracted lst with rbind
combined_lst <- extracted_lst %>% 
  rbind(time_lst)

#Found function t to convert rows to columns
fixed_combined <- t(combined_lst)

#Converted data with time to dataframe
fixed_combined <- as.data.frame(fixed_combined)

#Renamed automatic column names given
fixed_combined <- fixed_combined %>%
  rename(year = V2) %>%
  rename(temp = V1)

#Converted dates to years
fixed_combined <- fixed_combined %>%
  mutate(year = as.Date(year))

#Used "span" to fit the trend better with loess method
ggplot(fixed_combined, aes(x = year, y = temp)) + geom_point() + geom_smooth(method = "loess", formula = "y ~ x", span = 0.01) + labs(title = "Time Series of LST Near Buffalo. NY", x = "Date", y = "Land Surface Temperature (Degrees C)")

```
Part 2: Monthly Climatology
Hint
Summarize the weekly data into monthly means using tapp().

💡 Hints:

Use index = "months" and fun = mean in tapp().
Rename layers to month names. You could use month.name[...] or find another way to do it.
Visualize with a ggplot using tidyterra::geom_spatraster and facet_wrap(~lyr) to compare months.
```{r}
#Used tapp to sort lst by month, had to include na.rm =TRUE to include all data
monthly_lst <- tapp(lst, fun=mean, index="months", na.rm=TRUE)

names(monthly_lst)

#Renamed layers to month names
names(monthly_lst) <- c(month.name)

summary(monthly_lst)

#Checked to make sure data carried through
terra::values(monthly_lst)

ggplot(monthly_lst) + geom_spatraster(data = monthly_lst) + stat_spatraster(data = monthly_lst) + facet_wrap(~lyr) + scale_fill_gradientn(colors = c("blue", "gray", "red"))

```


Part 3: Compare LST by Land Cover Type

Explore how LST differs between Urban & built-up and Deciduous Broadleaf Forest areas.

💡 Hints:

Resample the land cover raster to match LST resolution with resample(method = "near").
Extract values from both rasters into a combined data frame. You could use bind_cols() or other methods.
Join land cover names from the MODIS legend table.
Filter to the two land cover types of interest (c("Urban & built-up", "Deciduous Broadleaf forest")) and plot monthly distributions.


```{r}

#I only chose data for one year as I was having trouble doing this for all year and monthly averages; I also used the monthly_lst rather than the general lst

lulc13 <- resample(lulc$Land_Cover_Type_1_13, monthly_lst, method = "near")

#Combined rasters
lulc_lst <- c(lulc13, monthly_lst)

#Converted combined rasters to dataset
lulc_lst <- as.data.frame(lulc_lst)

#Changed name to combine with lcd
lulc_lst <- lulc_lst %>%
  rename(ID = Land_Cover_Type_1_13)

#Formatted as.character to ensure proper joining
lulc_lst_filtered <- lulc_lst %>%
  mutate(ID = as.character(ID))

lulc_lst_filtered <- lulc_lst_filtered %>% 
  filter(ID == c("4", "13"))

#Formatted as.character to ensure proper joining
lcd <- lcd %>%
  mutate(ID = as.character(ID))

lulc_lst_filtered <- lulc_lst_filtered %>% left_join(lcd, by = "ID")

#Changed individual month columns to one column
lulc_lst_filtered <- lulc_lst_filtered %>%
  pivot_longer(cols = "January":"December", names_to = "month", values_to = "temperature")

#Formatted month column so it would go in month order on graph
lulc_lst_filtered$month <- factor(lulc_lst_filtered$month, levels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))
  
ggplot(data = lulc_lst_filtered, aes(x = month, y = temperature, fill = landcover)) + geom_violin() + labs(title = "Monthly LST For Landcover Type", x = "Month", y = "LST (Degrees C)", fill = "Landcover Type")
```

#Reflection
1. In summer, LST reaches its peak, and it takes a bit to decline going into colder months. But, LST rises more quickly going into summer months. There is also more temperature variation on average in the summer.
2. Forested areas stay cooler compared to urban areas, and there is less variation in temperature, especially in summer months.
3. I think MODIS is useful, but I found combining and separating LST for each month at different scales to be difficult and not very straightforward.

